Plant trait cleaning and combining

### 2024-06-10 code/L1/L1_P4_TropicalAndes_plant_traits.Rmd
add code to detect and report outliers, as well as summarize species records

---
title: "Plant trait cleaning and combining"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script subsets plant trait data by synonyms species list and combine."
data input: "ADD"
data output: "ADD"
date: "2023-11-04"
output: html_document
---

# Load required packages
```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(funbiogeo)
library(visdat)
```

# Set file paths
```{r}
data_path_L0<-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L0')
data_path_L1 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
figure_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/figures')
```

@@

Final trait list:
- dispersal syndrome
- Fruit mass
- Fruit length
- Plant growth form
- Plant height
- Plant Lifespan
- Seed mass
- Seed length
- Seed width

@@

  # Calculate average value, count, and variance per trait per species
  summary_data <- data %>%
    group_by(Accepted_species) %>%
    summarise(
      TraitValue_mean = mean(TraitValue, na.rm = TRUE),
      records_used = n(),
      variance = var(TraitValue, na.rm = TRUE)
    )

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

# Combine all traits
```{r}
trait_dfs <- list(averagetraits_fruitLength, averagetraits_fruitMass, averagetraits_plantHeight, averagetraits_plantLifespan, averagetraits_seedLength, averagetraits_seedMass, averagetraits_seedWidth, averagetraits_dispersal, averagetraits_growthForm)

@@

```{r}
# Assign data types
wide_plant_traits$`FruitLength_mm` <- as.numeric(wide_plant_traits$`FruitLength_mm`)
wide_plant_traits$`FruitMass_mg` <- as.numeric(wide_plant_traits$`FruitMass_mg`)
wide_plant_traits$`PlantHeight_m` <- as.numeric(wide_plant_traits$`PlantHeight_m`)
wide_plant_traits$`PlantLifespan_years` <- as.numeric(wide_plant_traits$`PlantLifespan_years`)
wide_plant_traits$`SeedLength_mm` <- as.numeric(wide_plant_traits$`SeedLength_mm`)
wide_plant_traits$`SeedMass_g` <- as.numeric(wide_plant_traits$`SeedMass_g`)
wide_plant_traits$`SeedWidth_mm` <- as.numeric(wide_plant_traits$`SeedWidth_mm`)
wide_plant_traits$`DispersalSyndrome` <- as.factor(wide_plant_traits$`DispersalSyndrome`)
wide_plant_traits$`GrowthForm` <- as.factor(wide_plant_traits$`GrowthForm`)
```

>>>>

---
title: "Plant trait cleaning and combining"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script subsets plant trait data by synonyms species list and combine."
data input: "TropicalAndes_all_plant_traits_harmonized_subset.csv"
data output: "ADD"
date: "2023-11-04"
output: html_document
---

# Load required packages
```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(rstatix)
library(funbiogeo)
library(visdat)
```

# Set file paths
```{r}
data_path_L1 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
figure_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/figures')
```

@@

Final trait list:
- dispersal syndrome
- Fruit mass
- Fruit length
- Plant growth form
- Plant height
- Plant Lifespan
- Seed mass
- Seed length
- Seed width
- Fruit type

@@

  # Calculate geometic mean value, count, and variance per trait per species
  summary_data <- data %>%
    group_by(Accepted_species) %>%
    summarise(
      TraitValue_mean = exp(mean(log(TraitValue))),
      records_used = n(),
      variance = var(TraitValue, na.rm = TRUE)
    )

@@

```{r}
# Detect outliers and report species with multiple entries per trait
detect_outliers <- function(data) {
  outlier_results <- data %>%
    group_by(Accepted_species) %>%
    filter(n() >= 3) %>%
    nest() %>%
    mutate(outliers = map(data, ~identify_outliers(.x, variable = "TraitValue"))) %>%
    unnest(outliers) %>%
    select(Accepted_species, TraitName, TraitValue, is.outlier, is.extreme) %>%
    ungroup()

  return(outlier_results)
}

# Get the species with outliers
species_with_outliers <- function(data) {
  outliers_by_species <- data %>%
    group_by(Accepted_species) %>%
    summarize(outlier_count = sum(is.outlier, na.rm = TRUE)) %>%
    filter(outlier_count > 0) %>%
    ungroup()

  return(outliers_by_species)
}
```

```{r}
summarize_species_records <- function(df) {
  # Count the total number of species
  total_species <- df %>% 
    distinct(Accepted_species) %>% 
    nrow()

  # Count the number of species with only one trait record
  species_with_one_record <- df %>% 
    group_by(Accepted_species) %>% 
    summarize(record_count = n()) %>% 
    filter(record_count == 1) %>% 
    nrow()

  # Count the number of species with more than three records
  species_with_more_than_three_records <- df %>% 
    group_by(Accepted_species) %>% 
    summarize(record_count = n()) %>% 
    filter(record_count > 3) %>% 
    nrow()

  # Print the results
  cat("Total number of species:", total_species, "\n")
  cat("Number of species with only one trait record:", species_with_one_record, "\n")
  cat("Number of species with more than three records:", species_with_more_than_three_records, "\n")
}
```

@@

```{r}
summarize_species_records(traits_dispersal)
```

@@

```{r}
dotchart(traits_fruitMass$TraitValue)
```

```{r}
# detect outliers
fruitMass_outliers <- detect_outliers(traits_fruitMass)
print(fruitMass_outliers)
fruitMass_outliers_species <- species_with_outliers(fruitMass_outliers)
print(fruitMass_outliers_species)
```

```{r}
summarize_species_records(traits_fruitMass)
```

@@

```{r}
dotchart(traits_fruitLength$TraitValue)
```

```{r}
# detect outliers
fruitLength_outliers <- detect_outliers(traits_fruitLength)
print(fruitLength_outliers)
fruitLength_outliers_species <- species_with_outliers(fruitLength_outliers)
print(fruitLength_outliers_species)
```

```{r}
summarize_species_records(traits_fruitLength)
```

@@

```{r}
dotchart(traits_plantHeight$TraitValue)
```

```{r}
# detect outliers
plantHeight_outliers <- detect_outliers(traits_plantHeight)
print(plantHeight_outliers)
plantHeight_outliers_species <- species_with_outliers(plantHeight_outliers)
print(plantHeight_outliers_species)
```

```{r}
summarize_species_records(traits_plantHeight)
```

@@

```{r}
dotchart(traits_plantLifespan$TraitValue)
```

```{r}
# detect outliers
plantLifespan_outliers <- detect_outliers(traits_plantLifespan)
print(plantLifespan_outliers)
plantLifespan_outliers_species <- species_with_outliers(plantLifespan_outliers)
print(plantLifespan_outliers_species)
```

```{r}
summarize_species_records(traits_plantLifespan)
```

@@

```{r}
dotchart(traits_seedMass$TraitValue)
```

```{r}
# detect outliers
seedMass_outliers <- detect_outliers(traits_seedMass)
print(seedMass_outliers)
seedMass_outliers_species <- species_with_outliers(seedMass_outliers)
print(seedMass_outliers_species)
```

```{r}
summarize_species_records(traits_seedMass)
```

@@

```{r}
dotchart(traits_seedLength$TraitValue)
```

```{r}
# detect outliers
seedLength_outliers <- detect_outliers(traits_seedLength)
print(seedLength_outliers)
seedLength_outliers_species <- species_with_outliers(seedLength_outliers)
print(seedLength_outliers_species)
```

```{r}
summarize_species_records(traits_seedLength)
```

@@

```{r}
summary(averagetraits_seedLength)
```

@@

```{r}
dotchart(traits_seedWidth$TraitValue)
```

```{r}
# detect outliers
seedWidth_outliers <- detect_outliers(traits_seedWidth)
print(seedWidth_outliers)
seedWidth_outliers_species <- species_with_outliers(seedWidth_outliers)
print(seedWidth_outliers_species)
```

```{r}
summarize_species_records(traits_seedWidth)
```

@@

```{r}
summarize_species_records(traits_growthForm)
```

@@

# Fruit type
```{r}
traits_fruitType <- plant_traits %>%
  filter(grepl('Fruit type|fruit type|fruit_type', TraitName))
traits_fruitType$OrigTraitValue <- traits_fruitType$TraitValue

nrow(traits_fruitType)
length(unique(traits_fruitType$Accepted_species))
```

```{r}
# remove numeric values
traits_fruitType <- traits_fruitType %>%
  filter(is.na(as.numeric(TraitValue)))

nrow(traits_fruitType)
length(unique(traits_fruitType$Accepted_species))
```

```{r}
traits_fruitType %>%
  count(TraitValue)
```

```{r}
# standardize trait values
traits_fruitType <- traits_fruitType %>%
  filter() %>%
  mutate(TraitValue= case_when(
    grepl('Berry|berry|Berry ', OrigTraitValue, ignore.case = TRUE) ~ "berry",
    grepl('capsule|Capsule|Capsule |Pixidium|Dehiscent capsule|Indehiscent capsule', OrigTraitValue, ignore.case = TRUE) ~ "capsule",
    grepl('drupe|Drupe|Drupe |Drupaceous', OrigTraitValue, ignore.case = TRUE) ~ "drupe",
    grepl('aggregate berries', OrigTraitValue, ignore.case = TRUE) ~ "aggregate berries",
    grepl('aggregate drupelets', OrigTraitValue, ignore.case = TRUE) ~ "aggregate drupelets",
    grepl('fleshy', OrigTraitValue, ignore.case = TRUE) ~ "fleshy",
    grepl('pome', OrigTraitValue, ignore.case = TRUE) ~ "pome",
    grepl('Squash', OrigTraitValue, ignore.case = TRUE) ~ "squash",
    grepl('aggregate follicles', OrigTraitValue, ignore.case = TRUE) ~ "aggregate follicles",
    grepl('aggregate nutlets', OrigTraitValue, ignore.case = TRUE) ~ "aggregate nutlets",
    grepl('achene', OrigTraitValue, ignore.case = TRUE) ~ "achene",
    grepl('follicle|Follicle ', OrigTraitValue, ignore.case = TRUE) ~ "follicle",
    grepl('legume|Legume ', OrigTraitValue, ignore.case = TRUE) ~ "legume",
    grepl('lomentum', OrigTraitValue, ignore.case = TRUE) ~ "lomentum",
    grepl('nut', OrigTraitValue, ignore.case = TRUE) ~ "nut",
    grepl('pod|Pod|Pod ', OrigTraitValue, ignore.case = TRUE) ~ "pod",
    grepl('Samara|Samaroid', OrigTraitValue, ignore.case = TRUE) ~ "samara",
    grepl('schizocarp', OrigTraitValue, ignore.case = TRUE) ~ "schizocarp",
    grepl('siliqua', OrigTraitValue, ignore.case = TRUE) ~ "siliqua",
    grepl('syncarpous', OrigTraitValue, ignore.case = TRUE) ~ "syncarpous",
    grepl('utricle', OrigTraitValue, ignore.case = TRUE) ~ "utricle",
    grepl('other|apocarpous|pericarp, aril, arillode|pseudosyncarpous|vegetative|dry|fleshy', OrigTraitValue, ignore.case = TRUE) ~ "unspecified",
  )) %>%
  filter(!is.na(TraitValue))
```

```{r}
traits_fruitType %>%
  count(TraitValue)
```

```{r}
# standardize trait name
traits_fruitType <- traits_fruitType %>%
  mutate(TraitName = "FruitType")
```

```{r}
summarize_species_records(traits_fruitType)
```

```{r}
averagetraits_fruitType <- factor_data_merge(traits_fruitType, "FruitType")
```

```{r}
nrow(averagetraits_fruitType)
length(unique(averagetraits_fruitType$Accepted_species))
```

```{r}
summary(averagetraits_fruitType)
```

# Combine all traits
```{r}
trait_dfs <- list(averagetraits_fruitLength, averagetraits_fruitMass, averagetraits_plantHeight, averagetraits_plantLifespan, averagetraits_seedLength, averagetraits_seedMass, averagetraits_seedWidth, averagetraits_dispersal, averagetraits_growthForm, averagetraits_fruitType)

@@

```{r}
# Assign data types
wide_plant_traits$`DispersalSyndrome` <- as.factor(wide_plant_traits$`DispersalSyndrome`)
wide_plant_traits$`FruitLength_mm` <- as.numeric(wide_plant_traits$`FruitLength_mm`)
wide_plant_traits$`FruitMass_mg` <- as.numeric(wide_plant_traits$`FruitMass_mg`)
wide_plant_traits$`FruitType` <- as.factor(wide_plant_traits$`FruitType`)
wide_plant_traits$`GrowthForm` <- as.factor(wide_plant_traits$`GrowthForm`)
wide_plant_traits$`PlantHeight_m` <- as.numeric(wide_plant_traits$`PlantHeight_m`)
wide_plant_traits$`PlantLifespan_years` <- as.numeric(wide_plant_traits$`PlantLifespan_years`)
wide_plant_traits$`SeedLength_mm` <- as.numeric(wide_plant_traits$`SeedLength_mm`)
wide_plant_traits$`SeedMass_g` <- as.numeric(wide_plant_traits$`SeedMass_g`)
wide_plant_traits$`SeedWidth_mm` <- as.numeric(wide_plant_traits$`SeedWidth_mm`)
```
